Erosion of somatic tissue identity with loss of the X-linked intellectual disability factor KDM5C

### 2024-08-20 submission/manuscript.Rmd
combine discussion cpgme and impact

# Discussion
In the above study, we demonstrate KDM5C's pivotal role in the development of tissue identity. We observed global dysregulation of tissue-enriched genes in the _Kdm5c_-KO brain and EpiLCs, including substantial misexpression of testis, liver, muscle, and ovary-enriched genes. The _Kdm5c_-KO amygdala and hippocampus had significant enrichment of testis genes that are specific to germ cells, rather than somatic cells. We found _Kdm5c_-KO EpiLCs aberrantly expressed key drivers of germline identity and meiosis, including _Dazl_ and _Stra8_, while the _Kdm5c_-KO brain primarily expressed germline genes important for late spermatogenesis. We demonstrated female EpiLCs are more sensitive to germline gene de-repression, as the number germline genes misexpressed and their degree of dysregulation was greater in _Kdm5c_-KO females compared to males. KDM5C is enriched at germline gene promoters in EpiLCs but is bound to only a subset of germline genes transcribed in _Kdm5c_-KO cells, indicating germline-enriched mRNAs can be aberrantly expressed through indirect mechanisms. Finally, we found KDM5C promotes the long-term silencing of germline genes in somatic cells by aiding the placement of DNA methylation at CpG islands for a subset of germline genes in EpiLCs. Therefore, we propose KDM5C plays a fundamental role in the development of tissue identity during early embryogenesis, including the establishment of the soma-germline boundary. By systematically characterizing KDM5C's role in germline gene repression during early embryogenesis, including its interaction with known silencing mechanisms, we unveiled unique regulatory mechansisms governing distinct germline gene classes in somatic lineages. Altogether, these results provide molecular footholds that in the future can be exploited to test the contribution of ectopic germline genes upon neurodevelopment.

@@

Our analysis of KDM5C ChIP-seq data provided novel insights into direct and indirect mechanisms by which different classes of germline genes can be misexpressed in _Kdm5c_-KO cells. Transcription of germline genes in the mature _Kdm5c_-KO brain is likely due to loss of KDM5c-mediated repression during embryogenesis, given that KDM5C was bound to their promoters in EpiLCs, is not recruited to germline genes in neurons, and germline transcription is not repressed when KDM5C in reintroduced in neurons[@scandagliaLossKdm5cCauses2017], 
It is unclear how KDM5C is recruited to germline genes, given that KDM5C itself does not contain domains for sequence-specific binding. In HeLa cells and ESCs[@tahilianiHistoneH3K4Demethylase2007; @guptaGenomewideScreenReveals2023], KDM5C assocaiates with members of the polycomb repressive complex 1.6 (PRC1.6), which is recruited to germline gene promoters through E2F6/DP1 and MGA/MAX heterodimers[@endohPCGF6PRC1SuppressesPremature2017; @stielowMGAL3MBTL2E2F62018]. While MAX and E2F6 motifs were enriched at KDM5C-bound germline promoters in EpiLCs, only about one-third of promtoers contained their consesus sequence. Additionally, germline genes like _Stra8_ can become active in _Kdm5c_-KO cells independent of KDM5C binding to their promoters. _Stra8_ is a meiotic transcription factor that is typically turned on in the presence of retinoic acid signaling and DAZL[@endoPeriodicRetinoicAcid2015; @bowlesRetinoidSignalingDetermines2006], which is a direct KDM5C target. Germline genes unbound by KDM5C were primarily late-stage spermatogenesis genes and significantly enriched for RFX transcription factors that promote the transcription of cilia and flagellar genes[@swobodaRFXTypeTranscriptionFactor2000; @ashiqueRfx4TranscriptionFactor2009]. One of the top RFX members enriched included RFX2, a central regulator of spermiogenesis[@kistlerRFX2MajorTranscriptional2015; @wuTranscriptionFactorRFX22016] whose mRNA is also expressed in _Kdm5c_-KO EpiLCs. Altogether, these data suggest that once germline genes are activated due to loss of direct KDM5C repression, they can activate other ectopic germline programs downstream that loosely mimic germline development.

### Germline gene decommissioning dna methylation
- End with comment on catalytic dependent vs independent functions of chromatin regulators
	- Requirement for catalytic activity may change depending upon the context

_Dazl_ and other germline-enriched genes are initially decommissioned in ESCs by repressive histone modifications and then _de novo_ DNA CpG methylation (CpGme) in EpiLCs[@velascoDnmt3bRecruitmentE2F62010; @karimiDNAMethylationSETDB12011; @hackettPromoterDNAMethylation2012; @tatsumiDNMTsSETDB1Function2018; @liuPolycombGroupProtein2020; @mochizukiRepressionGermlineGenes2021]. Unlike the previously characterized germline gene suppressors, KDM5C removes an active histone mark, histone 3 lysine 4 di- and trimethylation (H3K4me2/3)[@iwaseXLinkedMentalRetardation2007]. In this study, we observed a global increase in H3K4me2/3 in _Kdm5c_-KO EpiLCs and amygdala around the transcription start site (TSS) of germline-enriched genes. However, KDM5C's catalytic activity may not be required for germline gene silencing, as it was recently found to be dispensible for repressing _Dazl_ in ESCs [@guptaGenomewideScreenReveals2023]. Although not necessary in ESCs, KDM5C's catalytic activity be required for CpGme in EpiLCs, as H3K4me3 impedes _de novo_ CpGme placement[@otaniStructuralBasisRecognition2009; @guoStructuralInsightAutoinhibition2015]. This is supported by previous work in the _Kdm5C_-KO adult hippocampus, which found CpGme is significantly eroded at at least two germline promoters[@scandagliaLossKdm5cCauses2017]. To elucidate the mechanism behind KDM5C-mediated silencing of germline genes, we characterized KDM5C in ESCs and EpiLCs and found its expression is dynamically regulated during differenitation. Furthermore, KDM5C acts during the transition from na&#239;ve to primed pluripotency to promote the initial placement of CpGme at CpG islands for the long-term silencing of germline genes.

- CPG islands greatly detrmine KDM5C recruitment - KDM5C is previously known to be enriched at CGIs. CGIs not typcially methylated, germline genes exception. KDM5C promotes this process.

### Functional consequences 
In addition to testis genes identified previously[@scandagliaLossKdm5cCauses2017], we found significant enrichment of muscle, liver, and even ovary-biased genes aberrantly expressed within the male _Kdm5c_-KO brain. Given the majority of tissue-enriched DEGs have no known brain functions, it is currently unknown if they impair _Kdm5c_-KO neurodevelopment and contribute to MRXSCJ-related phenotypes like intellectual disability and aberrant aggression. However, select liver and muscle-biased DEGs do have known roles within the brain, such as the liver-enriched lipid metabolism gene _Apolipoprotein C-I (Apoc1)_ [@roulandRoleApolipoproteinC12022] that is highly expressed in the _Kdm5c_-KO amygdala and hippocampus. Intriguingly, _Apoc1_ overexpression in the mouse brain can impair learning and memory[@abildayevaHumanApolipoproteinCI2008] and is implicated in Alzheimer's disease in humans[@leducAPOECholesterolHomeostasis2010], however further investigation is required to determine its impact upon _Kdm5c_-KO phenotypes.

The above work provides the mechanistic foundation for KDM5C's role in germline gene repression and offers novel insight into how dysregulation in _Kdm5c_-KO tissue identity changes over the course of development. However, the contribution of ectopic, tissue-specific genes towards _Kdm5c_-KO neurological impairments is still unknown. KDM5C may be crucial for neurodevelopment by fine-tuning the expression of tissue-erniched, dosage-sensitive genes, such as _Apoc1_. KDM5C could also ensure tissue-specific timing of transcription factor family members, like RFX genes that are broadly required for cilia and flagella formation, including neuronal cilia[@swobodaRFXTypeTranscriptionFactor2000; @ashiqueRfx4TranscriptionFactor2009; @chungRFX2BroadlyRequired2012; @coschieraPrimaryCiliaPromote2024]. While their impact upon neurodevelopment is currently unclear, ectopic germline transcripts are also found in models of other related neurodevelopmental disorders[@bonefasSomatogermlineTransformationChromatinlinked2021], including Immunodeficiency, Centromere instability, Facial anomalies syndrome 1 (ICF1 OMIM: #242860)[@velascoGermlineGenesHypomethylation2014; @waltonDnmt3bPrefersGerm2014], Kleefstra syndrome 1 (OMIM: #610253)[@schaeferControlCognitionAdaptive2009], and MeCP2 duplication syndrome (MDS, OMIM: #300260)[@samacoCrhOprm1Mediate2012]. Like KDM5C, the chromatin regulators underlying these conditions, DNA methyltransferase 3b (DNMT3B), H3K9me1/2 methyltransferases G9A/GLP, methyl-CpG -binding protein 2 (MECP2), primarily silence gene expression. Thus, KDM5C is among a growing cohort of chromatin-linked neurodevelopmental disorders that have a similar underlying phenotype of germline versus soma dysregulation. Ultimately, further research is required to determine the impact of these germline genes and the extent to which this phenomenon occurs in humans.

---

Deeper characterization of testis-enriched DEGs revealed they were not somatic testis genes but instead germline genes, thus demonstrating KDM5C's crucial role in establishing the soma-germline boundary. Distinguishing the germline from the soma is a key feature of multicellularity and sexual reproduction[@devlinPanmetazoanViewGermlinesoma2023]. Previous work characterizing repressors of germ cell-specific transcription has predominatly focused on their repression of key marker genes in embryonic stem cells (ESCs), such as _Dazl_ and _Ddx4_[@endohPCGF6PRC1SuppressesPremature2017; @dahletE2F6InitiatesStable2021; @mochizukiRepressionGermlineGenes2021]. To characterize KDM5C's role in germline gene repression throughout life at a genome-wide level, we curated a list of germline-enriched genes using publically available germ cell-depleted RNA-seq datasets from Kit^W/Wv^ mice[@muellerIndependentSpecializationHuman2013; @sohGeneRegulatoryProgram2015]. This resource enabled us to identify 1) the extent of germline gene dysregulation in _Kdm5c_-KO cells, 2) the types of germline genes misexpressed at different developmental time points, 3) the impact of germ cell sex upon germline gene misexpression, and 4) which groups of germline genes are directly and indirectly regulated by KDM5C. This list of germline-enriched genes serves as a resource for future studies to systematically assess soma-germline dysregulation.

Analysis of epiblast-like cells (EpiLCs) revealed dysregulation of _Kdm5c_-KO tissue identity begins during early emrbyogenesis. _In vivo_, germline genes are typically fully decommissioned in epiblast stem cells and remain silenced as the epiblast differentiates into somatic tissues[@bardotMouseGastrulationCoordination2020]. However, a small subset of epiblast stem cells will reactivate germline genes to become the primoridal germ cells (PGCs) that will ultimately form the mature germline[@magnusdottirHowMakePrimordial2014; @gunesdoganPrimordialGermCell2014]. This process can be mimicked _in vitro_ by differentiating EpiLCs into primordial germ cell-like cells (PGCLCs)[@hayashiReconstitutionMouseGerm2011].

>>>>

# Discussion
In the above study, we demonstrate KDM5C's pivotal role in the development of tissue identity. We observed global dysregulation of tissue-enriched genes in the _Kdm5c_-KO brain and EpiLCs, including substantial misexpression of testis, liver, muscle, and ovary-enriched genes. The _Kdm5c_-KO amygdala and hippocampus had significant enrichment of testis genes that are specific to germ cells, rather than somatic cells. We found _Kdm5c_-KO EpiLCs aberrantly expressed key drivers of germline identity and meiosis, including _Dazl_ and _Stra8_, while the _Kdm5c_-KO brain primarily expressed germline genes important for late spermatogenesis. We demonstrated female EpiLCs are more sensitive to germline gene de-repression, as the number germline genes misexpressed and their degree of dysregulation was greater in _Kdm5c_-KO females compared to males. KDM5C is enriched at germline gene promoters in EpiLCs but is bound to only a subset of germline genes transcribed in _Kdm5c_-KO cells, indicating germline-enriched mRNAs can be aberrantly expressed through indirect mechanisms. Finally, we found KDM5C is dynamically regulated during the differentiation of ESCs in EpiLCs and promotes long-term germline gene silencing by aiding the placement of DNA methylation at CpG islands in EpiLCs. Therefore, we propose KDM5C plays a fundamental role in the development of tissue identity during early embryogenesis, including the establishment of the soma-germline boundary. By systematically characterizing KDM5C's role in germline gene repression during early embryogenesis, including its interaction with known silencing mechanisms, we unveiled unique regulatory mechansisms governing distinct germline gene classes in somatic lineages. Altogether, these results provide molecular footholds that in the future can be exploited to test the contribution of ectopic germline genes upon neurodevelopment.

@@

Our analysis of KDM5C ChIP-seq data provided novel insights into direct and indirect mechanisms by which different classes of germline genes can be misexpressed in _Kdm5c_-KO cells. Transcription of germline genes in the mature _Kdm5c_-KO brain is likely due to loss of KDM5c-mediated repression during embryogenesis, given that KDM5C was bound to their promoters in EpiLCs, is not recruited to germline genes in neurons, and germline transcription is not repressed when KDM5C in reintroduced in neurons[@scandagliaLossKdm5cCauses2017]. It is unclear how KDM5C is recruited to germline genes, given that KDM5C itself does not contain domains for sequence-specific binding. In HeLa cells and ESCs[@tahilianiHistoneH3K4Demethylase2007; @guptaGenomewideScreenReveals2023], KDM5C assocaiates with members of the polycomb repressive complex 1.6 (PRC1.6), which is recruited to germline gene promoters through E2F6/DP1 and MGA/MAX heterodimers[@endohPCGF6PRC1SuppressesPremature2017; @stielowMGAL3MBTL2E2F62018]. While MAX and E2F6 motifs were enriched at KDM5C-bound germline promoters in EpiLCs, only about one-third of promtoers contained their consesus sequence. Additionally, germline genes like _Stra8_ can become active in _Kdm5c_-KO cells independent of KDM5C binding to their promoters. _Stra8_ is a meiotic transcription factor that is typically turned on in the presence of retinoic acid signaling and DAZL[@endoPeriodicRetinoicAcid2015; @bowlesRetinoidSignalingDetermines2006], which is a direct KDM5C target. Germline genes unbound by KDM5C were primarily late-stage spermatogenesis genes and significantly enriched for RFX transcription factors that promote the transcription of cilia and flagellar genes[@swobodaRFXTypeTranscriptionFactor2000; @ashiqueRfx4TranscriptionFactor2009]. One of the top RFX members enriched included RFX2, a central regulator of spermiogenesis[@kistlerRFX2MajorTranscriptional2015; @wuTranscriptionFactorRFX22016] whose mRNA is also expressed in _Kdm5c_-KO EpiLCs. Altogether, these data suggest that once germline genes are activated due to loss of direct KDM5C repression, they can activate other ectopic germline programs downstream that loosely mimic germline development.

Unlike the previously characterized germline gene suppressors[@karimiDNAMethylationSETDB12011; @hackettPromoterDNAMethylation2012; @auclairEHMT2DirectsDNA2016; @endohPCGF6PRC1SuppressesPremature2017; @tatsumiDNMTsSETDB1Function2018; @liuPolycombGroupProtein2020; @mochizukiRepressionGermlineGenes2021], KDM5C removes an active histone mark, histone 3 lysine 4 di- and trimethylation (H3K4me2/3)[@iwaseXLinkedMentalRetardation2007]. However, KDM5C's catalytic activity may not be required for germline gene silencing, as it was recently found to be dispensible for repressing _Dazl_ in ESCs [@guptaGenomewideScreenReveals2023]. Emerging work indicate many histone-modifying enzymes have non-cataltyic functions that can influnce gene expression at least as potently as their catalytic roles[@aubertUnexpectedNoncatalyticRoles2019; @morganReevaluatingRolesHistonemodifying2020]. While dispensible in ESCs, KDM5C's catalytic activity may be necessary in EpiLCs, as H3K4me3 impedes _de novo_ CpGme placement[@otaniStructuralBasisRecognition2009; @guoStructuralInsightAutoinhibition2015] and CpGme is significantly eroded at at least two germline promoters in the adult _Kdm5c_-KO hippocampus[@scandagliaLossKdm5cCauses2017]. Indeed, we found _Kdm5c_-KO EpiLCs have impaired CpGme accumulation at germline gene CpG islands, indicating H3K4me3 removal is necessary at this developmental stage. Thus, the shift in the mechanism of KDM5C-mediated repression during the transition from na&#239;ve to primed pluripotency demonstrates the cataltyic activity of histone-modifying enzymes can become necessary within different cellular contexts.

- DNAme and CpG islands
	- Loss of many chromatin regulators affects CpGme (Multiple, non-redundant silencing mechansisms) - seems to be the crux point of germline gene silencing
	- Germline genes are methylated at promoter CGIs, which are typically not methylated in other gene types
	- Because CGIs are typically resistant to CpGme (accurate?), germline CGIs may require a highly repressive histone landscape to recruit sufficient DNMTs to these loci
	- We found not all germline-enriched genes have CGIs, but still gain CpGme around TSS. Unclear what the recruitment mechanism is for these genes
	- Other studies on germline gene repressors have shown they are importnat for CGIme, unclear if they also participate in non-CGI TSS CpGme
	- CGIs highly determine KDM5C recruitment to germline gene promoters and KDM5C germline gene CGI methylation.
		- KDM5C is known to be enriched at CGIs (in neurons? are these methylated? or is its germline CGI function likely different from its somatic CGI function). 
	- Different classes of germline genes use distinct silencing mechanisms

The above work provides the mechanistic foundation for KDM5C's role in germline gene repression and offers novel insight into how dysregulation in _Kdm5c_-KO tissue identity changes over the course of development. However, the contribution of ectopic, tissue-specific genes towards _Kdm5c_-KO neurological impairments is still unknown. In addition to germline genes, we also found significant enrichment of muscle, liver, and even ovary-biased genes aberrantly expressed within the male _Kdm5c_-KO brain. Intriguingly, select liver and muscle-biased DEGs do have known roles within the brain, such as the liver-enriched lipid metabolism gene _Apolipoprotein C-I (Apoc1)_ [@roulandRoleApolipoproteinC12022] that is highly expressed in the _Kdm5c_-KO amygdala and hippocampus. _Apoc1_ overexpression in the mouse brain can impair learning and memory[@abildayevaHumanApolipoproteinCI2008] and is implicated in Alzheimer's disease in humans[@leducAPOECholesterolHomeostasis2010]. Thus, KDM5C may be crucial for neurodevelopment by fine-tuning the expression of tissue-erniched, dosage-sensitive genes like _Apoc1_. While the impact of ectopic germline transcripts upon neurodevelopment is currently unknown, some models of other chromatin-linked neurodevelopmental disorders also display impaired soma-germline demarcation[@bonefasSomatogermlineTransformationChromatinlinked2021; @velascoGermlineGenesHypomethylation2014; @waltonDnmt3bPrefersGerm2014; @schaeferControlCognitionAdaptive2009; @samacoCrhOprm1Mediate2012]. Like KDM5C, the chromatin regulators underlying these conditions - DNA methyltransferase 3b (DNMT3B), H3K9me1/2 methyltransferases G9A/GLP, methyl-CpG -binding protein 2 (MECP2) - primarily silence gene expression. Thus, KDM5C is among a growing cohort of chromatin-linked neurodevelopmental disorders that have a similar underlying phenotype of germline versus soma dysregulation, however further research is required to determine the impact of these germline genes and the extent to which this phenomenon occurs in humans.
